When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-12-28 doc/paper/vpd_et_paper.tex
need to discuss vpd = 3.5 for pet in plots

Changes to vapor pressure deficit (VPD) alter the atmospheric demand for water from the land surface. Traditionally, atmospheric scientists and hydrometeorologists generally think that an increase in atmospheric demand induces an increases in evapotranspiration (ET) (citations?). This possible misconception developed in part due to the proliferation of studies examing potential ET (PET) rather than estimates of ET itself (citations?).  In contrast, plant physiologists know that stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase (decrease) in VPD may not correspond to an increase (decrease) in ET because stomatal closure (opening) can cancel the effects of shifts to atmospheric demand.

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved. An increase in ET in reponse to atmospheric drying will reduce soil moisture, but contribute increased moisenting to the atmosperhere. Clearly, the sign and magnitude of land-surface responsedrives the co-evolution of the atmosphere and land-surface at many timescales, from diurna to interdecadal.

We hypothesize that for most plant types a common resonse to increase in VPD will actually be a decrease in ET. The excpetion would be plants such as crops that are evolved (or bred) to prioritize gross primary production (GPP) over water conservation. However, for all otehr plants types, our hypothesis calls into question the validity of PET-based drought metrics deceloped by hydrometeorologists and used extensively in operations \citep[e.g. PDSI, P-PET,][]{2002, Alley_1984} \explain{more citations needed, including recent PET climate studies like Jack Scheff}. These metrics ignore the role of plants as gatekeepers for surface water loss to the atmosphere and have limited physical meaning for drought of vegetated land types. Additionally, plants evolved in arid climates should prioritize water conservation and we would expect a very negative ET response to increase in VPD. Therefore, vegetated locations most likely to experience droughts should show the strongest deviation between reality and a PET-based approximation.

@@

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes $\sigma$'s histogram.***}
\label{lai_vpd_fig}
\end{figure}

@@

ENF is a special case. The VPD dependence of the theory is not well captured by observations. However, the theory closely captures the intercept where $\frac{\partial \; ET}{\partial \; VPD} = 0$. Really resolving this feature is the key to answering our research question, so perhaps the theory is still useful for ENF for the purpose of addressing when VPD drives or reduces ET. It is also noteworthy that we have used published constants for $g1$, which determines the VPD dependence, from \cite{Lin_2015}. Had we fit our own constants we could have tweaked the theory to match the observations, but this was not our goal.

As stated in the above discussion of ENF, we are mostly interested in the critical threshold $VPD_{crit}$ for which the sign of  $\frac{\partial \; ET}{\partial \; VPD}$ is dertmined. When $VPD < VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}  < 0$, and when $VPD > VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD} > 0$. Theoretically this $VPD_{crit}$ is only a function of PFT. We can look at how this value of $VPD_{crit}$ matches the data as different observed variables (including uncertainty) are allowed to vary.

@@

The full observations generally confirm expectations from Section \ref{results}. CRO has the most positive values of $\frac{\partial \; ET}{\partial \; VPD}$, $\frac{\partial \; ET}{\partial \; VPD}$ is almost always negative for CSH, and response depends more with the environmental conditions for the other PFTs (especially ENF). Through the columns of Figure \ref{real} we can see the impact of $\sigma$ and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; VPD}$. $g_a$'s scaling (included in columns 1 and 3) alters the magnitude considerably, particularly begtween observations. $\sigma$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; VPD}$. However, the general story with the noise appears to match the cleaner signal when $\sigma$ is help constant and $VPD_{crit}$ is clearly visible. The exceptions are, as discussed above, for GRA and CRO, wher the full complexity plots (Columns 1 and 2) to not match well with $\sigma$ held fixed (Columns 3 and 4).

While $g_a$ has a large impact on magnitudes in the scaling, temperature, through the $\Delta$ term, does not appear to have as much of an impact in the full observations. Therefore it appears that the absolute magnitude of the derivative can be mostly attributed to the sign term, and its direct scaling by aerodynamic conductance. Intuitively this makes sense; aerodynamic conductance will control how dominant balances at the land surface are communicated to the boundary layer.

In general the full complexity plots of $\frac{\partial \; ET}{\partial \; VPD}$ match our expectations, even with the large sensitivity to $\sigma$ measures of uncertainty observed in Figure \ref{term3}. Our $\sigma$-based method of uncertainty propagation blurs the idealized expectations the most for GRA, and also has a considerable effect for CRO. We therefor have the most confidence in our conclusion based on Equation \ref{d_et} for PFTS CSH, DBF, and ENF, as the full complexity plots with uncertainty included closely match the story when $\sigma$ is held fixed.

>>>>

Answering the question ``When does vapor pressure deficit drive or reduce evapotranspiration?'' is motivated by two potenential but opposing perspectives on the matter. The first persepctive is that increases to vapor pressure deficit (VPD) increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indicies and hydrometeorological studies, only quantifies this atmospheric demand perspective. However, there is another possible perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD may actually correspond to a decrease in ET because stomatal closure can cancel the effects of shifts to atmospheric de. So really a question of ``When does VPD drive or reduce ET?'' becomes a question of ``When does plant response or atmospheric demand dominate?.''

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved. An increase in ET in reponse to atmospheric drying will reduce soil moisture, but contribute increased moisenting to the atmosperhere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

Before diving into math and data, we have some prior intuition for how the system might behave. First, whether VPD drive or reduces ET should be a function of plant type. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Additionally, atmospheric conditions must matter as well. At the ecosystem scale there are limits to the strategies plants use to hold water back from the atmosphere. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes $\sigma$'s histogram.***}
\label{lai_vpd_fig}
\end{figure}nnn

@@

ENF is a special case. The VPD dependence of the theory is not well captured by observations. However, the theory closely captures the intercept where $\frac{\partial \; ET}{\partial \; VPD} = 0$. Really resolving this feature is the key to answering our research question, so perhaps the theory is still useful for ENF for the purpose of addressing when VPD drives or reduces ET. It is also noteworthy that we used published constants for $g1$, which determines the VPD dependence, from \cite{Lin_2015}. We could have fit our own constants to tweak the theory to match the observations, but this was not our goal.

As stated in the above discussion of ENF, we are mostly interested in the critical threshold $VPD_{crit}$ for which the sign of  $\frac{\partial \; ET}{\partial \; VPD}$ is determined. When $VPD < VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}  < 0$, and when $VPD > VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD} > 0$. Theoretically this $VPD_{crit}$ is only a function of PFT. We can look at how this value of $VPD_{crit}$ matches the data as different observed variables (including uncertainty) are allowed to vary.

@@

The scatter plots generally confirm expectations from Figure \ref{agu_fig}. By comparing different columns of Figure \ref{real} we can see the impact of $\sigma$ and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; VPD}$. $g_a$'s scaling (included in columns 1 and 3) alters the magnitude considerably, particularly begtween observations. $\sigma$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; VPD}$, as observed in Figure \ref{agu_fig}. However, the general story with the noise appears to match the cleaner signal when $\sigma$ is help constant and $VPD_{crit}$ is clearly visible. The exceptions are, as discussed above, for GRA and CRO, where the full complexity plots (Columns 1 and 2) to not match well with $\sigma$ held fixed (Columns 3 and 4).

While $g_a$ has a large impact on magnitudes in the scaling, temperature, through the $\Delta$ term, does not appear to have as much of an impact in the full observations. Therefore it appears that the absolute magnitude of the derivative can be mostly attributed to the sign term, and its direct scaling by aerodynamic conductance. Intuitively this makes sense; aerodynamic conductance will control how dominant balances at the land surface are communicated to the boundary layer. While $\Delta$ controlls the efficiency of latent fluxes in carying heat away from the surface, it appears this is a second order term, relative to $g_a$, for scaling $\frac{\partial \; ET}{\partial \; VPD}$.

In general Figures \ref{real} and \ref{agu_fig} match our expectations, even with the large sensitivity to $\sigma$ seen in Figure \ref{term3}. Our $\sigma$-based method of uncertainty propagation blurs the idealized expectations the most for GRA and CRO. We therefor have the most confidence in our conclusion based on Equation \ref{d_et} for PFTS CSH, DBF, and ENF, as the sign of the derivative with uncertainty included matches (to leading order) the story when $\sigma$ is held fixed.
